Harmony: A Systems Programming Language and Model Checker

### 2022-01-18 paper/paper.tex

\emph{Harmony} is a Python-like programming language designed to teach
systems specification, implementation, and verification techniques to
students who do not have a background in formal methods.
Harmony programs, which can be multi-threaded
or distributed, are model checked~\cite{CES86}.
Without any special formalism, Harmony can diagnose a large variety
of problems in programs:
\begin{compactitem}
\item ``Silly errors'': accessing non-existent variables,
divide-by-zero, out-of-bounds array access, dereferencing dangling
pointers, trying to multiply two lists with one another,
and many more.
\item Safety violations due to assertions failing, such as an assertion to ensure mutual exclusion.
\item Liveness violations such as caused by deadlock or infinite loops.
\item Faulty behaviors: implementation has behaviors not seen in the specification, such as violations of linearizability~\cite{HW90}.
\item Incomplete behaviors: specification has behaviors not seen with implementation, such as multiple readers not able to get a reader/writer lock simultaneously.
\item Data races: shared variables accessed without a lock.
\item Busy Waiting: waiting for a condition without blocking.
\item Violations of interrupt safety: a race condition caused by the execution
of an interrupt handler.
\item Memory leaks.
\end{compactitem}

@@

Harmony's unique combination of features are a familiar and easy
to learn high-level and highly expressive programming language, the
ability to concisely specify an interface, the ability to compare
behaviors of specification and implementation, the ability to
modularly model check programs, and support for dynamic memory
allocation, continuations, and interrupts.  In addition, Harmony
finds various common problems automatically without a complicated
property specification language, such as Harmony's ability to find
instances of undesirable busy waiting.

@@

Harmony does not
support classes and instead supports memory allocation and pointers
in a way that is familiar to C programmers.  While this perhaps
seems like a step back in language evolution, many synchronization
algorithms in the literature explicitly consider memory accesses,
and this has been a bonus for analyzing such algorithms with Harmony.
Moreover, the Harmony model checker can be used to check for memory
leaks or access to dangling pointers.

@@

\emph{Deadlock} is the situation in which threads are waiting for
one another.
In order to make deadlock detection possible, Harmony programs are
required to eventually terminate under fair selection of its
nondetermistic choices.  In other words, from each state there
should always exist an extension of the execution in which the
Harmony program terminates.  Harmony supports \emph{eternal}
threads---threads that never terminate---for example to model
servers.  A Harmony program is considered terminated if it is in a
state in which all its remaining threads are eternal and blocked.

@@

To test a queue, we want to perform a set of concurrent <{enqueue}>
and <{dequeue}> operations concurrently.
\autoref{fig:queue}(b) shows how this can be specified in Harmony.
The test program spawns <{NOPS = 4}> threads, at least one of which
does an <{enqueue}> operation and at least one of which does a
<{dequeue}> operation.  A thread that does an <{enqueue}> enqueues
its identifier <{self}>.  The program
includes a <{print}> statement at the beginning and end of every queue
operation.  Think of the <{print}> statements as \emph{annotations}
that tell Harmony what the interesting events are with respect to
queue operations.

@@

The test program is similar to what a programmer might have written
as a unit test.  The programmer would run the program a bunch of
times and look at the output in order to see if it looks reasonable.
Doing so is difficult and error prone, as the output tends
to be complex and in the mess of many lines of output it is easy
to overlook problems such as two threads dequeuing the same item
or items being dequeued in the wrong order.  The output of concurrent
print statement can be mangled together.  Not only that, but the
programmer would need to run this test program many times and check
the output each time.  It even is hard to write a program that
checks the output automatically.

@@

\begin{compactitem}
\item \emph{basic operations}: do a simple operation to a context
such as pushing a value onto its stack, and increase its program
counter by one.
\item \emph{jumps}: to implement loops.
\item \texttt{Choose}: nondetermistic selection from a set.
\item \texttt{Apply}: to call a method \emph{or} to index into a dictionary.
\item \texttt{Frame/Return}: method start/stop.
\item \texttt{Spawn}: create a new thread.
\item \texttt{AtomicInc/AtomicDec}: modify the atomic counter of the context.
\item \texttt{SetIntLevel}: enable/disable interrupts.
\item \texttt{Stop/Go}: manipulating continuations.
\item \texttt{Assert}: check a value.
\item \texttt{Print}: add a value to the print log.
\end{compactitem}

@@

The little orange boxes represent individual HVM instructions of the
problematic trace.  You can click on any such box to see the state
at that time and the HVM instruction it is executing,
or you can navigate in various ways (forwards one step,
backwards one step, jump to the next line of Harmony code, or jumps
forwards or backwards to a context switch between threads).  The actual
HVM code is in the lower left of the output.

@@

PlusCal~\cite{Lamport09} is an imperative programming language that,
after translation to TLA+~\cite{Lamport02} (a formal verification
language), can be run through the TLC model checker.  Harmony also
supports compilation to TLA+.  While similar in spirit to Harmony,
Harmony first compiles to a virtual machine that is close to CPUs
like x86, ARM, and RISC-V, using a stack and common control flow
techniques.  The resulting HVM has a TLA+ specification.  PlusCal,
instead, \emph{trans\-piles} to TLA+.  Informally, the gap between
PlusCal programs and their TLA+ representations is much smaller
than the gap between Harmony programs and their TLA+ representations,
making PlusCal programs more amenable to formal verification.  On the other
hand, working effectively with PlusCal does require significant
knowledge of TLA+, increasing the learning curve.

>>>>

\emph{Harmony} is a Python-like programming language designed to teach
systems specification, implementation, and verification techniques to
students who do not have a background in formal methods.
Harmony programs, which can be multi-threaded
or distributed, are model checked~\cite{CES86}.
Without any special formalism, Harmony can diagnose a large variety
of problems in programs:
\begin{compactitem}
\item ``Silly errors'': accessing non-existent variables,
divide-by-zero, out-of-bounds array access, dereferencing dangling
pointers, trying to multiply two lists with one another,
and many more.
\item Safety violations: Harmony supports both assertions that must hold in a particular location and general invariants.
\item Liveness violations: not reaching a final state, often due to deadlock or an infinite loop.
\item Faulty behaviors: implementation has behaviors not seen in the specification, such as violations of linearizability~\cite{HW90}.
\item Incomplete behaviors: specification has behaviors not seen with implementation, such as multiple readers not able to get a reader/writer lock simultaneously.
\item Data races: shared variables accessed without a lock.
\item Busy Waiting: waiting for a condition without blocking.
\item Violations of interrupt safety: a race condition caused by the execution
of an interrupt handler.
\item Memory leaks: memory unintentionally allocated in a final state.
\end{compactitem}

@@

Harmony's unique combination of features are
\begin{compactitem}
\item a familiar and easy-to-learn high-level programming language.
\item the ability to concisely specify an interface.
\item the ability to compare behaviors of specification and implementation.
\item the ability to modularly model check programs.
\item support for dynamic memory allocation.
\item support for continuations.
\item support for interrupts.
\item automatic diagnosis of various common problems such as
undesirable busy waiting.
\end{compactitem}

@@

Harmony does not
support classes and instead supports memory allocation and pointers
in a way that is familiar to C programmers.  While perhaps this
seems like a step back in language evolution, many synchronization
algorithms in the literature explicitly consider memory accesses,
and this has been a bonus for analyzing such algorithms with Harmony.
Moreover, the Harmony model checker can be used to check for memory
leaks or access to dangling pointers.

@@

\emph{Deadlock} is the situation in which threads are waiting for
one another.
In order to make deadlock detection possible, a Harmony program is
required to eventually terminate under fair selection of its
nondetermistic choices.  In other words, from each state there
should always exist an extension of the execution in which the
Harmony program terminates.  Harmony supports \emph{eternal}
threads---threads that never terminate---for example to model
servers.  A Harmony program is considered terminated if it is in a
state in which all its remaining threads are eternal and blocked.

@@

To test a concurrent queue, we need to perform a set of <{enqueue}>
and <{dequeue}> operations concurrently.  \autoref{fig:queue}(b)
shows how this can be specified in Harmony.  The test program spawns
<{NOPS = 4}> threads, at least one of which does an <{enqueue}>
operation and at least one of which does a <{dequeue}> operation.
A thread that does an <{enqueue}> enqueues its identifier <{self}>.
The program includes a <{print}> statement at the beginning and end
of every queue operation.  Think of the <{print}> statements as
\emph{annotations} that tell Harmony what the interesting events
are with respect to queue operations.  We will use the output of
those <{print}> statements to compare the behavior of an implementation
with that of its specification.

@@

The test program is similar to what a programmer might have written
as a unit test.  The programmer would run the program a bunch of
times and look at the output in order to see if it looks reasonable.
Doing so is difficult and error prone, as the output tends to be
complex and---in the mess of many lines of output---it is easy to
overlook problems such as two threads dequeuing the same item or
items being dequeued in the wrong order.  Not only that, but the
programmer would need to run this test program many times and check
the output each time.  It is hard to write a program that checks
the output automatically as it is difficult to characterize what
correct outputs are.

@@

\begin{compactitem}
\item \emph{basic operations}: a deterministic operation that only
involves a single context and no shared state, such as pushing a value
onto the stack or increment the program counter.
\item \texttt{Frame/Return}: method start/end.
\item \texttt{Choose}: nondetermistic selection from a set.
\item \texttt{Apply}: method call \emph{or} index into a dictionary.
\item \texttt{Spawn}: thread creation.
\item \texttt{AtomicInc/AtomicDec}: atomic section enter/leave.
\item \texttt{SetIntLevel}: interrupt enable/disable.
\item \texttt{Stop/Go}: continuation save/restart.
\item \texttt{Assert}: value check.
\item \texttt{Print}: output.
\end{compactitem}

@@

The little orange boxes represent individual HVM instructions of
the problematic trace.  The programmer can click on any such box
to see the state at that time and the HVM instruction it is executing,
or can navigate in various ways (forwards one step, backwards
one step, jump to the next line of Harmony code, or jumps forwards
or backwards to a context switch between threads).  The actual HVM
code is in the lower left of the output.

@@

PlusCal~\cite{Lamport09} is an imperative programming language that,
after translation to TLA+~\cite{Lamport02} (a formal verification
language), can be run through the TLC model checker.  Harmony also
supports compilation to TLA+.  While similar in spirit to Harmony,
Harmony first compiles to a virtual machine that is close to CPUs
like x86, ARM, and RISC-V, using a stack and common control flow
techniques.  The resulting HVM has a TLA+ specification.  PlusCal,
instead, \emph{trans\-piles} to TLA+ (source-to-source translation).
Informally, the gap between PlusCal programs and their TLA+
representations is much smaller than the gap between Harmony programs
and their TLA+ representations, making PlusCal programs more amenable
to formal verification.  On the other hand, working effectively
with PlusCal does require significant knowledge of TLA+, increasing
the learning curve.
